Modelling writing hesitations in text writing as finite mixture process

### 2023-09-15 report/manuscript.Rmd
Add simulation

---
title             : "Modelling writing hesitations in text writing as finite mixture process"
shorttitle        : "Modelling writing hesitations"

@@

abstract: |

@@

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE
    includes:
      after_body: 
        - "appendix_a.tex"
        - "appendix_b.tex"
        - "appendix_c.tex"
        - "appendix_d.tex"
        - "appendix_e.tex"
        - "appendix_g.tex"
        - "appendix_f.tex"

@@

beta_label <- "Short key transition\nduration (in msecs)"
delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
theta_label <- "Probability of hesitant\ntransitions"
```

# Introduction

@@

@@

The models presented in the following can be divided into two general groups. The first three models are largely akin to models typically used in the literature. By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models. Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies. The latter two models map directly on the idea of a cascading model of writing.

### Unimodal Gaussian

We start with a Gaussian mixed-effects models similar to a standard analysis of variance. We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$. The mean can be decomposed into $\beta$ and $\text{participant}_i$. $\beta$ will be allowed to take on a different value for each transition location. Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.

\begin{equation}
\begin{aligned}
(\#eq:unimodgaus)
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Unimodal log-Gaussian

This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution. The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero. For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key. Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses. In particular, a normal distribution assumes that units are linearly scaled. For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e. 5,050 msecs). This is not necessarily plausible for keystroke data. We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2). Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).

The model can be described like this:

\begin{equation}
\begin{aligned}
(\#eq:unimodloggaus)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Unimodal unequal-variance log-Gaussian

\begin{equation}
\begin{aligned}
(\#eq:unimoduv)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Bimodal log-Gaussian (constrained)

This models extends the intuition from the previous model that higher levels of activation lead to longer pauses. Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two. (1) activation can flow into keystrokes without interrupts. These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$. In principle, there are no differences for fluent key-transitions between transition location. The next model with loosen this assumption (hence, "unconstrained model"). (2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time. The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word). Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive). (2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

```{r}
#		<!-- &\delta \sim \text{N}(0,1)\\ -->

This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

### Bimodal log-Gaussian (unconstrained)

This model is identical to the previous model with one exception. The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model. In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions. This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.

In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.

\begin{equation}
\begin{aligned}
(\#eq:bimoduncon)
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

```{r}
#<!-- & \delta \sim \text{N}(0,1)\\ -->

@@

Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.

>>>>

---
title             : "Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process"
shorttitle        : "Modelling writing hesitations"

@@

abstract: Classical serial models view the writing process as a chain of pauses and writing burst. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations. Our results cannot be explain by the serial but are in line with the cascading view of writing.

@@

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE
    includes:
      after_body: 
        - "appendix_a.tex"
        - "appendix_b.tex"
        - "appendix_c.tex"
        - "appendix_d.tex"
        - "appendix_e.tex"
        - "appendix_g.tex"
        - "appendix_f.tex"
        - "appendix_g.tex"

@@

beta_label <- "Short transition\nduration (in msecs)"
delta_label <- "Slowdown for hesitant\ntransitions (in secs)"
theta_label <- "Probability of hesitant\ntransitions"
```

Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].

Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech. Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language. Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written]. This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g. @Clark2002]. Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].

The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models. On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06]. For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses]. The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01]. We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted. Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.

<!-- a statistical framework for the analysis of keystroke data that captures not just the multilevel-structure of the data but also presents a principled way to capture the probabilistic nature of pauses in text composition. Hesitations are probabilistic in the sense that we cannot be predicted their occurrence deterministically. In principle we can produce several words and even sentences without pausing. However, hesitations are more of less likely to occur in certain contexts, for example, at sentence boundaries.  -->

Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward]. The intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured. Processing may be serial. This means that the processing cascade operates on one unit at a time. Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing]. This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].

This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second]. However this account is not consistent with what we know about writing. Consider these two observations:

First, utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process. This is well known for spoken production [e.g. @bock2014syntactically]. For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance]. For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e. *A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset. This is not what one would expect if writers plan sentences (or even phrases) in full before production onset. In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e. *the B*) was more likely for conjoined noun phrases.

Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid. For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration. Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control. Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction]. @van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now]. Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two data-generating processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e. just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].

For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling. In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach. Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions. As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes. For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.

The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. @ronneberg2022process coined this possibility the "process-disruption hypothesis".

In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].

Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production. Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating;@dux2021effect;@vandermeulen2023writing].

Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse]. Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016. These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].

Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual]. In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks. Instead text composition provides data from both inter and intra-sentence transitions. Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.

@@

We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as unimodal models. In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component [@wagenmakers2007linear; @wing1973response; @schoner2002timing].

The remaining two models are implementations of the cascading view following the bimodal mixture-model approach presented in @roeser2021modelling. Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow. Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur. This view was implemented as a constrained and an unconstrained model.

@@

We are using the Bayesian framework [e.g. @farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing. In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models. The models presented in this section build on one another so that later model include assumptions of earlier models.

### Serial model of writing

#### Unimodal Gaussian

Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit. The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors. For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphes, syllables, and morphemes. There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis. We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model. This can be expressed as $\text{iki}_\text{before word} \sim \text{N}(\mu, \sigma_\text{e}^2)$. Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.

We can extend this simple model of word-planning to other linguistic location. We introduce earlier that larger linguistic edges are associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07]. Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors. We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries. However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs. Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g. whether there is a statistically meaningful difference between IKI associated with words and sentences).

For computational ease, we implemented the differences associated with transition locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location. Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs. Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$. The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$. This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e. a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).

The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model.

\begin{equation}
\begin{aligned}
(\#eq:unimodgaus)
\text{iki}_i \sim\text{ } & \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

Note aside that standard-deviation parameters were constrained to be positive because standard deviations can never be negative.

#### Unimodal log-Gaussian

The previous model assumes that the data-generating process is a Gaussian distribution. The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution. There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. IKIs, as the distance between two subsequent key-down events, must be positive. The lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not necessarily map onto the psychological interpretation for short and long keyintervals. For example effects that result on the motor level within words (e.g. typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).

The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{N}()$ was replaced by $\text{logN}()$.

\begin{equation}
\begin{aligned}
(\#eq:unimodloggaus)
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
& u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

#### Unimodal unequal-variance log-Gaussian

The third model representing the serial view of writing is an unimodal unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section. The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text. The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing]. For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies as a function of transition location.

\begin{equation}
\begin{aligned}
(\#eq:unimoduv)
\text{iki}_i \sim\text{ } & \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + u_\text{participant[i]}\\
 & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Parallel cascading model of writing

The following two models are are extensions of the models introduced for the serial view. Crucially, the cascading view allows planning to happen in parallel to production. Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset. This is done by assuming that IKIs come from a weighted combination of two distributions.

#### Bimodal log-Gaussian (constrained)

This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses. Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory. This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:

1. Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}. Note that $\beta$ in both log-Gaussian distributions in equation \ref{eq:bimodcon} refers to the same unknown parameter.

2. Interruptions in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when words or their spelling could not be retrieved in time. The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}. The magnitude of the slowdown is associated with transition location. This is because delays at larger linguistic units are likely to be associated with higher level planning.  By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 > \theta > 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions. Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1. $\theta$ is here parametrised to represent the probability that an IKI is associated with the distribution of long IKIs. This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$. In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution. The weights of both distributions must sum to 1. We will call this parameter the probability of hesitant transitions.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less. This is akin to what is known as a random by-participant slopes model.

Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location. In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature]. This was achieved by introducing a parameter $\sigma_\text{diff}$. The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.

#### Bimodal log-Gaussian (unconstrained)

In principle, we do not need to assume that the size of a fluent key-transition varies by transition location. In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions. This is what we called a constrained model. However, letter bigrams (or trigrams) may be executed faster than transitions between between space and a letter (REFERECE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section). This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions outside of words. Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data. We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.

In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.

\begin{equation}
\begin{aligned}
(\#eq:bimoduncon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: }  & u_\text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

@@

Five datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.
